---
title: "20250720_finding_cnv_overlaps_v2"
author: "Tatjana"
format: html
editor: visual
---

## Computing CNV overlaps with GenomicRanges - version 2

after chatting w Kristy on 31.07, try out a new strategy for overlapping CNVs that were called using the "spot-based" CNV analysis.

```{r}
library(tidyverse)
library(magrittr)
library(qs)
library(GenomicRanges)
```

```{r}
# testing approach on a mini arbitrary sample to check how all functions behave
grtest <- GRanges(
    seqnames = Rle(c("chr1", "chr2", "chr1", "chr3"), c(1, 3, 2, 4)),
    ranges = IRanges(101:110, end = 111:120, names = head(letters, 10)),
    strand = Rle(strand(c("*")), c(10)),
    score = 1:10,
    GC = seq(1, 0, length=10))
grtest = grtest[1:6]
```

```{r}
djtest = disjoin(grtest)
```

```{r}
oltest = countOverlaps(djtest, grtest)
mcols(djtest)$count = oltest
```

```{r}
# make df of the data we are interested in
dftest = data.frame("width" = djtest@ranges@width, "count" = djtest$count)

# plot histogram of widths, weightes by counts
ggplot(data = dftest, 
       mapping = aes(x = width, weight = count)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Histogram of bin widths",
    x = "width of bin (bp)",
    y = "frequency"
  )

# looks as expected, now proceed with patint samplesß
```

```{r}
# test sample of 1 patient
test = read.table("/scratch/tomek/01_siCNV/RESULTS_spotbased_infercnv_per_roi/LN0040HD/LN0040HD_12BV2R_R1/HMM_CNV_predictions.HMMi6.hmm_mode-cells.Pnorm_0.5.pred_cnv_regions.dat", header = TRUE)
```

```{r}
test$width = test$end - test$start
test$width_kb = (test$width)/1000
test_subset_to_keep = grep(paste0("^", "LN0040HD", "_"), test$cell_group_name)
test_flt = test[test_subset_to_keep, , drop=FALSE]
test_flt_lst = split(test_flt, test_flt$chr)
```

```{r}
#gr = GRanges(seqnames = test_flt$chr,
             #ranges = IRanges(
               #start = test_flt$start,
               #end = test_flt$end,
               #names = test_flt$cnv_name),
              #barcode = test_flt$cell_group_name,
              #state = test_flt$state,
              #cnv_name = test_flt$cnv_name
             #)
```

```{r}
gr_list = lapply(test_flt_lst, function(bc) {
  GRanges(
    seqnames = bc$chr,
    ranges = IRanges(
      start = bc$start,
      end = bc$end, 
      names = bc$cnv_name),
    barcode = bc$cell_group_name,
    state = bc$state,
    cnv_name = bc$cnv_name
    )
  })
```

First, plot size distribution of CNVs found

```{r}
ggplot(test_flt_lst[[1]], aes(x = width_kb)) +
  geom_histogram(binwidth = 1000) +
  theme_minimal() +
  #scale_x_log10() +
  labs(
    title = "CNV sizes",
    x = "cnv size (kb)",
    y = "freq"
  )
```

```{r}
dj_list = lapply(gr_list, function(gr_obj) {
  dj = disjoin(gr_obj)
})

#dj = disjoin(gr)
```

Now: count how often each sequence chunk appears in "gr", by running "countOverlaps"

```{r}
ol_list = mapply(function(dj_obj, gr_obj) {
  overlap = countOverlaps(dj_obj, gr_obj)
  mcols(dj_obj)$count = overlap
  dj_obj
}, dj_list, gr_list)

#ol = countOverlaps(dj, gr)
#mcols(dj)$count = ol
```

Next, make df of data we want to plot

```{r}
df = data.frame(
  "width_kb" = (ol_list[[1]]@ranges@width)/1000,
  "count" = ol_list[[1]]$count)

df = df[order(df$count, decreasing = TRUE),]
```

```{r}
# make plot
ggplot(data = df,
       mapping = aes(x = width_kb, weight = count)) +
  geom_histogram(binwidth = 200) +
  #scale_x_continuous(trans='log10')
  theme_minimal()
```

```{r}
# make log10-transformed width colum; filter out zero or negative values
df$log_width_kb <- log10(df$width_kb[df$width_kb > 0])

ggplot(df, aes(x = log_width_kb, weight = count)) +
  geom_histogram(binwidth = 0.1) +  # tomatch logscale
  xlab("log10(width [kb])") +
  theme_minimal()

```

```{r}
ol_list_chr1 =  ol_list[[1]]
ol_list_chr1 = ol_list_chr1[order(ol_list_chr1@ranges@width)]
```

```{r}
ol_list_chr1 = ol_list_chr1[order(ol_list_chr1$count, decreasing = TRUE)]
```

```{r}
# looking at count distribution of counts found on chr1
counthist_chr1 = hist(df$count)
```

remove sequnces that were found in \< 200 cnvs

```{r}
ol_list_chr1 = ol_list_chr1[ol_list_chr1$count > 200]
df_trimmed = data.frame(
  "width_kb" = (ol_list_chr1@ranges@width)/1000,
  "count" = ol_list_chr1$count)
```

```{r}
counthist_chr1 = hist(df_trimmed$count)
```

```{r}
try_reduce = reduce(ol_list_chr1)
```

```{r}
try_reduce$cnv_state = c("del", "del", "del", "del")
```

```{r}
# make gtrellis plot
pdf("/g/saka/Tatjana/data/data_for_testing_stuff/LN0040HD_R1_cnv_test.pdf", width = 15, height = 4)

col_fun = c("amp" = "red", "del" = "blue")


gtrellis_layout(
  add_name_track = TRUE,
  n_track = 1, 
  equal_width = TRUE,
  title = "CNVs inferred using inferCNV - LN0040HD R1", 
  track_ylab = c("inferCNV"), 
  track_axis = FALSE,
  xlab = "")

add_rect_track(
  try_reduce,
  track = 2,
  h1 = 1, 
  h2 = 0,
  gp = gpar(col = NA, fill = col_fun[mcols(try_reduce)$cnv_state])
)

dev.off()

## it worked!! we could recover all cnvs that we expected for chromosome 1 in this patient sample - do for all samples & clean up
```
